---
title: "Using rerddapUtils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using rerddapUtils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(duckdb)
library(duckplyr)
library(lubridate)
library(ncdf4)
library(rerddap)
library(rerddapUtils)
```
 
 `rerddapUtils` is an [R]{style="color:red"} package is a set of four functions desigened to work with and extend the `rerddap` package to provide capabilities requested by users that meet specialized needs which are better not being included in the `rerddao` package.
 
 The first function is:
 
- `griddap_season <- function(datasetx, ..., fields = 'all', stride = 1, season = NULL, fmt = "nc", url = eurl(), store = disk(), read = TRUE, callopts = list())`
                          
which will only extract data for the period during the year defined by 'season'.  The second function is:
  
  
- `griddap_split <- function(datasetx, ..., fields = 'all', stride = 1, request_split = NULL, fmt = "nc", url = eurl(), store = disk(), read = TRUE, callopts = list(), aggregate_file = NULL)`
              
which is designed to split a very large request into pieces defined by 'request_split' and then aggregate the parts either in memory,  in a duckdb database or in a netcdf4 file.  This allows for requests larger than 2GB to be made in an appropriate manner.
 
Note that `griddap_season()` and `griddap_split()` are designed to have basically the same interface as `rerddao::griddap()`,  except for one or two  extra function arguments and some arguments that are interpted differently. In both functions the argument 'store = disk()' is ignored.  In `griddap_season` the 'fmt' argument is ignored  while 'season' defines the days during the year to make the extract.

In  `griddap_split` the argument 'fmt' defines whether the aggregate download should be stored in memory as a dataframe,  in a duckdb database file,  or in a netcdf file.  For the duckdb and nectdf options, 'aggregate_file' lets you define where to write the file,  if not defined it will set up a temporary file following proper [R]{style="color:red"} guidelines and in both cases the path to the file will be returned.

The other two functions are designed to make it easier to work with projected datasets:

- `latlon_to_xy <- function (dataInfo, latitude, longitude, yName = 'latitude', xName = 'longitude', crs = NULL) `

which for a given dataset will convert a latitude and longitude request into the projected coordiantes to be used in `rerddao::griddap()`, while the fourth function:

- `xy_to_latlon <- function (resp, yName = 'cols', xName = 'rows', crs = NULL)`

does the reverse,  given an `rerddao::griddap()` extract will convert the projected coordinates into latitiude and longitude values.

This vignette assumes familiarity with the various 'rerddap' functions and how to use them.  More information about the various 'rerddap' functions cna be found in the documentation for that package.

## Extract during a season - `griddap_season()`

The function `griddap_season()` works exactly like the function `rerddap::griddap` except that there is one extra argument,  "season",  which restricts the extract to the part of the year defined by season, and also ignores several of the `rerddap::griddap` options,  as it will ignore what format to downaod the data,  and will only save to a standard `rerddap::griddap` output structure.

The argument "season" is a character string such that for a given "year",  paste0(year, '-', season) produces a valid ISO datetime string.  To test "season" is properly formed, use `lubridate::as_datetime()`:

```{r}
year <- '2023'
season <- c('02-01', '06-01')
# test that "season" is defined properly
test_time <- paste0(year, '-', season)
lubridate::as_datetime(test_time)
```

Even though the example only includes month and day,  hours etc can be included as long as the final result,  when a year is appended,  produces a valid datetime.

As an example call and result, suppose you have a working `rerddap::griddap()` extract given by:

```{r}
wind_info <- rerddap::info('erdQMekm14day')
extract <- rerddap::griddap(wind_info,
                            time = c('2015-01-01','2019-01-01'),
                            latitude = c(20, 40),
                            longitude = c(220, 240),
                            fields = 'mod_current'
)
```

but it is desired to restrict extract to March 1 to May 1 inclusive,  then the following modification of the above using `griddap_season()`produces the desired result:


```{r}
wind_info <- rerddap::info('erdQMekm14day')
season <- c('03-01', '05-01')
season_extract <- griddap_season(wind_info,
 time = c('2015-01-01','2019-01-01'),
 latitude = c(20, 40),
 longitude = c(220, 240),
 fields = 'mod_current',
 season = season
)

```

To see that the extract has been limited to the given season:

```{r}
extract_times  <- lubridate::as_datetime(season_extract$data$time)
extract_months <- lubridate::month(extract_times)
unique(extract_months)
```

Only results for months 3, 4, and 5 are included,  as per "season".  May is included  because the season definition is inclusive,  hence May 1 will be included.

```{r}
#library(duckdb)
#library(duckplyr)

# Establish a connection to the DuckDB file
con <- duckdb::duckdb("test.duckdb")

# Assuming the DataFrame is stored in a table named 'extract'
# Use duckplyr to work with the table using dplyr syntax
extract_tbl <- duckplyr::tbl(con, "extract")

# You can then perform operations on the table. For example, to read and print the entire DataFrame:
extract_df <- duckplyr::collect(extract_tbl)

# Print the DataFrame
print(extract_df)

# Don't forget to disconnect when you're done
duckdb::dbDisconnect(con)

```

