---
title: "Using rerddapUtils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using rerddapUtils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(duckdb)
library(duckplyr)
library(lubridate)
library(ncdf4)
library(rerddap)
library(rerddapUtils)
```
 
 `rerddapUtils` is an [R]{style="color:red"} package is a set of four functions desigened to work with and extend the `rerddap` package to provide capabilities requested by users that meet specialized needs which are better not being included in the `rerddao` package.
 
 The first function is:
 
- `griddap_season <- function(datasetx, ..., fields = 'all', stride = 1, season = NULL, fmt = "nc", url = eurl(), store = disk(), read = TRUE, callopts = list())`
                          
which will only extract data for the period during the year defined by 'season'.  The second function is:
  
  
- `griddap_split <- function(datasetx, ..., fields = 'all', stride = 1, request_split = NULL, fmt = "nc", url = eurl(), store = disk(), read = TRUE, callopts = list(), aggregate_file = NULL)`
              
which is designed to split a very large request into pieces defined by 'request_split' and then aggregate the parts either in memory,  in a duckdb database or in a netcdf4 file.  This allows for requests larger than 2GB to be made in an appropriate manner.
 
Note that `griddap_season()` and `griddap_split()` are designed to have basically the same interface as `rerddao::griddap()`,  except for one or two  extra function arguments and some arguments that are interpted differently. In both functions the argument 'store = disk()' is ignored.  In `griddap_season` the 'fmt' argument is ignored  while 'season' defines the days during the year to make the extract.

In  `griddap_split` the argument 'fmt' defines whether the aggregate download should be stored in memory as a dataframe,  in a duckdb database file,  or in a netcdf file.  For the duckdb and nectdf options, 'aggregate_file' lets you define where to write the file,  if not defined it will set up a temporary file following proper [R]{style="color:red"} guidelines and in both cases the path to the file will be returned.

The other two functions are designed to make it easier to work with projected datasets:

- `latlon_to_xy <- function (dataInfo, latitude, longitude, yName = 'latitude', xName = 'longitude', crs = NULL) `

which for a given dataset will convert a latitude and longitude request into the projected coordiantes to be used in `rerddao::griddap()`, while the fourth function:

- `xy_to_latlon <- function (resp, yName = 'cols', xName = 'rows', crs = NULL)`

does the reverse,  given an `rerddao::griddap()` extract will convert the projected coordinates into latitiude and longitude values.

This vignette assumes familiarity with the various 'rerddap' functions and how to use them.  More information about the various 'rerddap' functions can be found in the documentation for that package.

## Extract during a season - `griddap_season()`

The function `griddap_season()` works exactly like the function `rerddap::griddap` except that there is one extra argument,  "season",  which restricts the extract to the part of the year defined by season, and also ignores several of the `rerddap::griddap` options,  as it will ignore what format to downaod the data,  and will only save to a standard `rerddap::griddap` output structure.

The argument "season" is a character string such that for a given "year",  paste0(year, '-', season) produces a valid ISO datetime string.  To test "season" is properly formed, use `lubridate::as_datetime()`:

```{r test_season_time}
year <- '2023'
season <- c('02-01', '06-01')
# test that "season" is defined properly
test_time <- paste0(year, '-', season)
lubridate::as_datetime(test_time)
```

Even though the example only includes month and day,  hours etc can be included as long as the final result,  when a year is appended,  produces a valid datetime.

As an example call and result, suppose you have a working `rerddap::griddap()` extract given by:

```{r wind_extract, eval=FALSE}
wind_info <- rerddap::info('erdQMekm14day')
extract <- rerddap::griddap(wind_info,
                            time = c('2015-01-01','2019-01-01'),
                            latitude = c(20, 40),
                            longitude = c(220, 240),
                            fields = 'mod_current'
)
```

but it is desired to restrict extract to March 1 to May 1 inclusive,  then the following modification of the above using `griddap_season()`produces the desired result:


```{r wind_season}
wind_info <- rerddap::info('erdQMekm14day')
season <- c('03-01', '05-01')
season_extract <- griddap_season(wind_info,
 time = c('2015-01-01','2019-01-01'),
 latitude = c(20, 40),
 longitude = c(220, 240),
 fields = 'mod_current',
 season = season
)

```

To see that the extract has been limited to the given season:

```{r season_months}
extract_times  <- lubridate::as_datetime(season_extract$data$time)
extract_months <- lubridate::month(extract_times)
unique(extract_months)
```

Only results for months 3, 4, and 5 are included,  as per the values of "season".  May is included  because the definition of "season" is inclusive,  hence May 1 will be included.

## Split an extract into parts - `griddap_split()`

The function `griddap_split()` allows to make `rerddap::griddap()` requests that are larger than the 2GB limit by splitting the request into parts and then combining the results either in memory in a dataframe,  in a netcdf4 file,  or into a duckdb database file.  If the "in memory" option is chosen, there is no check that there is adequate memory,  so it is possible to crash both R and your computer. The nectdf4 and duckdb options depend more on having adequate disk space,  not memory.  Also,  there is no check that the split given will reduce the requests to a small enough size, that is up to the user to estimate.

`griddap_split()` uses the same arguments as `griddap_split()` execpt that:

- The "store" and "read" arguments are ignored.
- "fmt" is one of "memory", "nc", or "duckdb" - default "nc"
- an added argument "split"
- an added argument "aggregate_file" - default is to create an R temporary file

The argument "split" is a named list that defines the number of splits in each dimension,  not the values where to make the splits, and a value must be given for each coordinate in the dataset.  So for the dataset 'erdQMekm14day' above,  there are four coordinates,  "time", "altitude", "latitude" and "longitude", to break the request into five "time" parts and two "latitude" and "longitude" parts  (if you remember your combinatorics the original request will be broken into 20 separate requests) use:

```{r request_split}
request_split <- list(time = 5, altitude = 1, latitude = 2, longitude = 2)
```

Note that:

```{r request_split_fail, eval = FALSE}
request_split <- list(time = 5, latitude = 2, longitude = 2)
```

will fail,  even though "altitude" is dimension one.  A value must be given for all of the dataset coordinates.

"fmt" defines how the split extracts will be combined to give one result.  If fmt = "memory"  then the results are combined into a usual `rerddap::griddap()`  in memory,  and that structure is returned by the function.  If fmt = "nc" then the extracts wil be combined into a netcdf4 file,  and a path to the netcdf file is returned by the function. If fmt = "duckdb" then the extracts wil be combined into a duckdb database file,  and a path to the duckdb file is returned by the function.

As in the previous example suppose I had an `rerddap::griddap()` extract:

```{r wind_extract_redux, eval=FALSE}
wind_info <- rerddap::info('erdQMekm14day')
extract <- rerddap::griddap(wind_info,
                            time = c('2015-01-01','2016-01-01'),
                            latitude = c(20, 40),
                            longitude = c(220, 240),
                            fields = 'mod_current'
)
```

which would request an extract too large for ERDDAP to fulfill (that is not the case here so a split is not needed,  but this illustrates the usage).  Then the following modifies the code to split the request and save into a structure in memory:

```{r winf_split_memory, eval = FALSE}

wind_info <- rerddap::info('erdQMekm14day')
request_split <- list(time = 5, altitude = 1, latitude = 2, longitude = 2)
split_extract <- griddap_split(wind_info,
 time = c('2015-01-01','2016-01-01'),
 latitude = c(20, 40),
 longitude = c(220, 240),
 fields = 'mod_current',
 request_split = request_split,
 fmt = "memory"
)
str(split_extract)
```

On return, "split_extract" will contain the usual `rerddap::griddap()` structure.  Again,  there is no check that there is enough memory to do this operation,  so it is possible to crash R (or your computer) if not used carefully,  but it is not unusual for computers to have 16GB of RAM or more,  so extracts a good bit larger than 2GB can be stored in memory.

If instead it is desired to store the result in a netcdf4 file:

```{r wind_split_nc, evsl = FALSE}

wind_info <- rerddap::info('erdQMekm14day')
request_split <- list(time = 5, altitude = 1, latitude = 2, longitude = 2)
split_extract <- griddap_split(wind_info,
 time = c('2015-01-01','2016-01-01'),
 latitude = c(20, 40),
 longitude = c(220, 240),
 fields = 'mod_current',
 request_split = request_split,
 fmt = "nc"
)

```

Since "aggregate_file" is not given, the netcdf file will be written to a temporary space and the path to that file will be returned by the function.  The reason for this is it is bad form for an R package to write to a user's space without permission.  To give permission to wrtie to a file of one's choosing, either give the full path to the file (or if just the name is given it will be written to the present directory), such as for the example above:

```{r wind_split_nc_named, eval=FALSE}

wind_info <- rerddap::info('erdQMekm14day')
request_split <- list(time = 5, altitude = 1, latitude = 2, longitude = 2)
split_extract <- griddap_split(wind_info,
 time = c('2015-01-01','2019-01-01'),
 latitude = c(20, 40),
 longitude = c(220, 240),
 fields = 'mod_current',
 request_split = request_split,
 fmt = "nc",
 aggregate_file = 'wind_extract.nc'
)

```

To access the data in the file, any of the various R packages that can read netcdf4 files can be used, in particular ones that allow for only part of the dataset to be read into memory at a time.  For example to open the result of the netcdf file created above using 'ncdf4' and viewing a summary:

```{r wind_split_read_nc, eval = FALSE}
wind_file <- ncdf4::nc_open(split_extract )
wind_file
ncdf4::nc_close(wind_file)

```


To make the same extract and save to a duckbb database file:

```{r wind_split_duckdb, eval = FALSE}

wind_info <- rerddap::info('erdQMekm14day')
request_split <- list(time = 5, altitude = 1, latitude = 2, longitude = 2)
split_extract <- griddap_split(wind_info,
 time = c('2015-01-01','2016-01-01'),
 latitude = c(20, 40),
 longitude = c(220, 240),
 fields = 'mod_current',
 request_split = request_split,
 fmt = "duckdb",
 aggregate_file = 'wind_extract.duckdb'
)

```


There are many tools that can used to access the duckdb database file,  among others the packages 'dplyr' and 'duckplyr' can be used:

```{r wind_split_read, eval = FALSE}
con <- dbConnect(duckdb::duckdb(), "wind_extract.duckdb")
tbl(con, "extract") |>
     head(5) |>
     collect()
dbDisconnect(con, shutdown=TRUE)
```

Note that the table will always be called "extract".


## Working with projected datasets - `latlon_to_xy()` and `xy_to_latlon()`


Extracting projected data in ERDDAP  (that is not on a latitude-longitude) grid can be difficult because most people do not think in terms of the projection in defining the region to make an extract,  and similary once an extract has been made,  it is often desirable to know the coordinates in terms of latitude and longitude.  The functions `latlon_to_xy()` and `xy_to_latlon()` are desigend to help with that process.

In both cases the functions try to determine the dataset projection either from the dataset summary given by `rerddap::info()` when projecting the values of a latitude-longitude bounding box,  and when an extract has been made, from the metadata in the extract.  This is done by searching for any of the terms:

```{r proj_strings, eval = FALSE}
  proj_strings <- c('proj4string', 'proj_crs_code', 'proj4text', 'grid_mapping_epsg_code',
                    'grid_mapping_proj4', 'grid_mapping_proj4_params', 'grid_mapping_proj4text',
                    'WKT')

```

in the metadata.  If this is not found,  then the user has to provide "CRS",  which can be in any of the forms above. For example the Ice dataset 'nsidcG02202v4sh1day' on the Polarwatch ERDDAP is projected. The bounding box of interest is:

```{r bounding_box}
latitude <- c(20, 30)
longitude <- c(-140, -130)

```

then the bounding box in projected coordinates is found from:

```{r latlon_to_xy, eval = FALSE}
myURL <- 'https://polarwatch.noaa.gov/erddap/'
myInfo <- rerddap::info('nsidcG02202v4sh1day', url = myURL)
latitude <- c(20, 30)
longitude <- c(-140, -130)
coords <- latlon_to_xy(myInfo, latitude, longitude)
coords

```

Similary,  given an extract on a projected coordinate system, the values on terms of latitude and longitude can be found by:

```{r xy_to_latlon, eval = FALSE} 
xgrid <- c(162500, 512500)
ygrid <- c(-437500,-637500)
xName <- 'xgrid'
yName <- 'ygrid'
proj_extract <- rerddap::griddap(myInfo, 
                time = c('2023-06-30T00:00:00Z', '2023-06-30T00:00:00Z'),
                ygrid = ygrid,
                xgrid = xgrid,
                fields = 'cdr_seaice_conc',
                url = myURL
                )
test <- xy_to_latlon(proj_extract, yName = 'ygrid', xName = 'xgrid')
head(test)

```


Note that `xy_to_latlon()` will also work with the output from the functions `rerddapXtracto::rxtracto()`, `rerddapXtracto::rxtracto_3D() ` and ``rerddapXtracto::rxtractogon()`.
